Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-15 examples_C.tex
+

\subsection{Mapping a New Page}
\label{sec:mapnew}
One of the key tasks of a page fault handler in a general-purpose OS kernel is
to map new pages into an address space by writing into an existing page table via a call\\
\centerline{\textsf{vaspace\_mappage(pte\_t *pml4, void *va,uintptr\_t fpaddr)}}\\
in Figure \ref{fig:mapping_code}.
To do so, with a given allocated a fresh page (\textsf{fpaddr}), then calculate the appropriate
known-valid page table walks (via \textsf{walkpgdir} Line \ref{line:call_walkpgdir} in Figure \ref{fig:mapping_code})  and update 
the appropriate L1 page table entry (Line 35 in Figure \ref{fig:mapping_code});
unmapping is the reverse of the logic we discuss here.
\looseness=-1
%\lstset{
%  columns=fullflexible,
%  numbers=left,
%  basicstyle=\ttfamily,
%  keywordstyle=\color{blue}\bfseries,
%  morekeywords={mov,add,call},
%  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
%  emphstyle=\color{green},
%  emph={[2]cr3},
%  emphstyle={[2]\color{violet}},
%  morecomment=[l]{;;},
%  mathescape
%}
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
/*
@param plm4: the root address of the page-table tree
@param va: virtual address to be translated and mapped
@param fpaddr: physically allocated page's address in case the virtual address va is not backed with one
*/
void vaspace_mappage(pte_t *pml4, void *va,uintptr_t fpaddr ) {
  $\specline{\textsf{P} \ast \textsf{pml4} \mapsto_{\textsf{id}} \textsf{\_} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \ulcorner\theta \; !!\; \vaddr = \texttt{None}\urcorner}_{\rtv}$      
  pte_t *pteaddr = walkpgdir(pml4, a); (*\label{line:call_walkpgdir}*)
  $\specline{ \textsf{R}_{\textsf{walk}} \ast \textsf{R}_{\textsf{l1e}} }$
  if (!pteaddr->present){ (*\label{line:mappage_pte_present_start}*)
    /*... ;; Not present(*\label{line:mappage_pte_present_end}*)*/
    pteaddr->pfn = PTE_ADDR_TO_PFN(fpaddr); /*Store updated entry back to L1 entry (*\label{line:l1entry_store}*)*/
    $\specline{ \textsf{pt\_entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{pt\_entry} \; (\textsf{set\_pfn}(\textsf{l1e\_val}, \textsf{PTE\_ADDR\_TO\_PFN}(\textsf{fpaddr}))) \ast  \ulcorner \lnot(\textsf{entry\_present l1e\_val}) \urcorner \ast \textsf{P}}$
    /*... set present bit in entry, then clean up the stack*/
    pteaddr->present = 1;
  }
  /*
  ... and return either unaccessible pte pointsto or one with fresh page (fpaddr)
  */
  return error;
  $\specline{ \begin{array}{l}\textsf{pt\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{pt\_entry} \; \textsf{l1e\_val} \ast \ulcorner (\textsf{entry\_present l1e\_val}) \urcorner \ast \textsf{P} \\
 \lor \; \;  \textsf{pt\_entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{pt\_entry} \; (\textsf{set\_pfn}(\textsf{l1e\_val}, \textsf{PTE\_ADDR\_TO\_PFN}(\textsf{fpaddr}))) \ast  \ulcorner \lnot(\textsf{entry\_present l1e\_val}) \urcorner \ast P \end{array} }$
}
\end{lstlisting}
\vspace{-1em}
  \caption{Specification of updating L1 entry (\textsf{pt\_entry}) to reference a new page (\textsf{fpaddr}).}
\label{fig:mapping_code}
\end{figure}

>>>>

\subsection{Mapping a New Page}
\label{sec:mapnew}
One of the key tasks of a page fault handler in a general-purpose OS kernel is
to map new pages into an address space by writing into an existing page table via a call\\
\centerline{\textsf{vaspace\_mappage(pte\_t *pml4, void *va,uintptr\_t fpaddr)}}\\
in Figure \ref{fig:mapping_code}.
To do so, with a given allocated a fresh page (\textsf{fpaddr}), then calculate the appropriate
known-valid page table walks (via \textsf{walkpgdir} Line \ref{line:call_walkpgdir} in Figure \ref{fig:mapping_code})  and update 
the appropriate L1 page table entry (Line 35 in Figure \ref{fig:mapping_code});
unmapping is the reverse of the logic we discuss here.
\looseness=-1
%\lstset{
%  columns=fullflexible,
%  numbers=left,
%  basicstyle=\ttfamily,
%  keywordstyle=\color{blue}\bfseries,
%  morekeywords={mov,add,call},
%  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
%  emphstyle=\color{green},
%  emph={[2]cr3},
%  emphstyle={[2]\color{violet}},
%  morecomment=[l]{;;},
%  mathescape
%}
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
/*
@param plm4: the root address of the page-table tree
@param va: virtual address to be translated and mapped
@param fpaddr: physically allocated page's address in case the virtual address va is not backed with one
*/
void vaspace_mappage(pte_t *pml4, void *va,uintptr_t fpaddr ) {
  $\specline{\textsf{P} \ast \textsf{pml4} \mapsto_{\textsf{id}} \textsf{\_} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \ulcorner\theta \; !!\; \vaddr = \texttt{None}\urcorner}_{\rtv}$      
  pte_t *pteaddr = walkpgdir(pml4, a); (*\label{line:call_walkpgdir}*)
  $\specline{ \textsf{R}_{\textsf{walk}} \ast \textsf{R}_{\textsf{l1e}} }$
  if (!pteaddr->present){ (*\label{line:mappage_pte_present_start}*)
    /*The entry is not present(*\label{line:mappage_pte_present_end}*)*/
    pteaddr->pfn = PTE_ADDR_TO_PFN(fpaddr); /*Store updated entry back to L1 entry (*\label{line:l1entry_store}*)*/
    $\specline{ \textsf{pt\_entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{pt\_entry} \; (\textsf{set\_pfn}(\textsf{l1e\_val}, \textsf{PTE\_ADDR\_TO\_PFN}(\textsf{fpaddr}))) \ast  \ulcorner \lnot(\textsf{entry\_present l1e\_val}) \urcorner \ast \textsf{P}}$
    /*Set present bit in entry, then clean up the stack*/
    pteaddr->present = 1;
  }
  return;
  $\specline{ \begin{array}{l}\textsf{pt\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{pt\_entry} \; \textsf{l1e\_val} \ast \ulcorner (\textsf{entry\_present l1e\_val}) \urcorner \ast \textsf{P} \\
 \lor \; \;  \textsf{pt\_entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{pt\_entry} \; (\textsf{set\_pfn}(\textsf{l1e\_val}, \textsf{PTE\_ADDR\_TO\_PFN}(\textsf{fpaddr}))) \ast  \ulcorner \lnot(\textsf{entry\_present l1e\_val}) \urcorner \ast P \end{array} }$
}
\end{lstlisting}
\vspace{-1em}
  \caption{Specification of updating L1 entry (\textsf{pt\_entry}) to reference a new page (\textsf{fpaddr}).}
\label{fig:mapping_code}
\end{figure}
